Abstractions for Multi-Sorted Substitutions

### 2023-09-16 paper/paper.tex
Update.

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservattion lemmas for renamings and substitutions.
  Section~\ref{sec:semantics} introduces a small-step operational semantics for System F.
  Section~\ref{sec:subjectreduction} proves the subject reduction lemma for System F.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:beyond} describes case studies of object languages beyond System F.
  Section~\ref{sec:future} describes future work.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:conclusion} concludes.

@@

  \section{Beyond System F}
  \label{sec:beyond}
  \begin{itemize}
  \item
    Binders with multiple variables are supported and even pattern
    matching can be encoded.
  \item
    Heterogenous equality
  \end{itemize}

>>>>

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservattion lemmas for renamings and substitutions.
  Section~\ref{sec:semantics} introduces a small-step operational semantics for System F.
  Section~\ref{sec:subjectreduction} proves the subject reduction lemma for System F.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} describes case studies of object languages beyond System F.
  Section~\ref{sec:future} describes future work.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:conclusion} concludes.

@@

  \section{Case Studies}
  \label{sec:casestudies}
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instanciating the
    \ACode{\ACon{TypingTraversal}}. Additionally, having the
    \ACode{\ACon{ComposeKit}} abstraction turned out to be useful
    to prove confluence, allowing to unify certain lemmas, which
    otherwise need to be proved for both renamings and substitutions,
    e.g. that for two terms \ACode{t} and \ACode{t'}
    with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    Instead of defining typing contexts and type preservation lemmas independent
    of our framework, we found a way that still allows us to benefit
    from the typing machinery.
    When binding a type variable with a subtyping constraint \ACode{α <: t},
    we first bind the type variable as \ACode{α : ★}, and then bind the constraint
    with a dummy name \ACode{\_ ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization.
    With that encoding substitution preserves typing is generally true again, enabling
    the use of the framework's typing contexts and \ACode{\ACon{TypingKit}} machinery.
    Replacing a type variable \ACode{α <: t} with a type \ACode{t'}
    which is not a subtype of \ACode{t'}, the resulting term is still
    well-typed, but in a context with an insatisfiable constraint.
  \item
    Object language with pattern matching pose an interesting problem.
    A clause \ACode{p ⇒ e} in a match-expression consists of a pattern
    \ACode{p} and a continuation expression \ACode{e}.
    The pattern \ACode{p} can bind arbitrary many variables in subpatterns,
    which are then available as free variables in \ACode{e}.
    % The challenge is to track the sorts of variables bound by a pattern,
    % to make them available to \ACode{e} by extending \ACode{e}'s sort
    % context correspondingly.
    To track the sorts of variables bound by a pattern, one can define
    the sort of patterns to contain the sorts of the variables that it binds:
    % 𝕡 : List (Sort Var) → Sort NoVar
    The syntax for variable patterns and clauses can then be specified as:
    % varpat : S ⊢ 𝕡 [ 𝕖 ]  -- pattern binding an expression variable
    % pairpat : S ⊢ 𝕡 S₁ → S ⊢ 𝕡 S₂ → S ⊢ 𝕡 (S₁ ++ S₂)
    % clause : S ⊢ 𝕡 S' → (S' ++ S) ⊢ 𝕖 → S ⊢ 𝕔
  \end{itemize}
